My wonderful paper

### 2024-11-07 _manuscript/_tex/index.tex
more on data analysis check literature

\citep{peng_diagnosing_2021}

@@

In data analysis, checks are often performed to ensure the quality of
the data and the validity of the model. For data checks,

\begin{itemize}
\tightlist
\item
  The \texttt{assertr} package \citep{assertr} provides a set of
  functions to check the quality of data, including the presence of
  missing values, the distribution of variables, and the relationship
  between variables.
\item
  Similarly, the \texttt{pointblank} package \citep{pointblank} is a
  data validation tool that allows users to define data quality checks.
\end{itemize}

For model checks,

\begin{itemize}
\tightlist
\item
  the \texttt{performance} package \citep{performance} provides a set of
  functions to assess the performance of a model, including residual
  analysis, goodness-of-fit tests.
\item
  The \texttt{yardstick} package \citep{yardstick} provides the
  calculation of model performance metrics in the \texttt{tidymodels}
  \citep{tidymodels} ecosystem.
\end{itemize}

These checks test on specific aspect of the data or the model but
doesn't reflect the underlying assumption of the data analysts about the
data or the problem context. It may be obvious to see the assumption if
someone is testing for data quality on whether a column in the data
frame is between two values. In a much complex analysis, it is difficult
to trace the assumptions made by the data analysts. Currently, most of
these assumptions are not documented or discussed in the analysis
process, or at least in a machine readable format.

>>>>

\citep{peng_diagnosing_2021} describes three pedagogical exercises of
introducing diagnosing unexpected outcome into a classroom setting.

@@

Currently, little has been done on how to computationally incorporate
this diagnostic approach into data analysis workflow or software. Most
of the diagnostic tools focus on defining user-specified rules, such as
data quality checks or producing metrics to summarize model performance,
as in model diagnostics. For example, the \texttt{assertr} package
\citep{assertr} and the \texttt{pointblank} package \citep{pointblank}
provide data validation tools to allow users to define data quality
checks. In contrast, packages provide model checks tools like
\texttt{performance} \citep{performance} and \texttt{yardstick}
\citep{yardstick}, from the \texttt{tidymodels} \citep{tidymodels}
ecosystem, offers goodness-of-fit, residual analysis, and performance
metrics.

These packages provide the tools to conduct diagnostics but still don't
reflect the mental process of how data analysts diagnose the unexpected
output. For example, when an unexpected output occurs, an analyst may
check on whether a column in the data frame is between two values for
data quality. However, it is not documented what motivates the analyst
to conduct this check -- whether it also applies to other researchers
analyzing new data in the same contexts, whether it is a common practice
in the field, or whether it is a reaction to this particular data or
scenario. Currently, most of these assumptions are not discussed in the
publication or captured by tools themselves. While one might be able to
infer some of the mental process of the analysts from external sources,
such as talking to them or watching screencast videos produced by the
analysts e.g.~TidyTuesday screencast videos, these insights are not
systematically documented or made machine-readable. This gap highlights
the need for tools that provide higher level documentation of reasoning
behind the checks, facilitating a more transparent and interpretable
analysis process.

### 2024-11-07 index.qmd
more on data analysis check literature

[@peng_diagnosing_2021]

@@

  In data analysis, checks are often performed to ensure the quality of the data and the validity of the model. For data checks,

  * The `assertr` package [@assertr] provides a set of functions to check the quality of data, including the presence of missing values, the distribution of variables, and the relationship between variables. 
  * Similarly, the `pointblank` package [@pointblank] is a data validation tool that allows users to define data quality checks.

  For model checks,

  * the `performance` package [@performance] provides a set of functions to assess the performance of a model, including residual analysis, goodness-of-fit tests. 
  * The `yardstick` package [@yardstick] provides the calculation of model performance metrics in the `tidymodels` [@tidymodels] ecosystem.

These checks test on specific aspect of the data or the model but doesn't reflect the underlying assumption of the data analysts about the data or the problem context. It may be obvious to see the assumption if someone is testing for data quality on whether a column in the data frame is between two values. In a much complex analysis, it is difficult to trace the assumptions made by the data analysts. Currently, most of these assumptions are not documented or discussed in the analysis process, or at least in a machine readable format.

>>>>

[@peng_diagnosing_2021] describes three pedagogical exercises of introducing diagnosing unexpected outcome into a classroom setting.

@@

Currently, little has been done on how to computationally incorporate this diagnostic approach into data analysis workflow or software. Most of the diagnostic tools focus on defining user-specified rules, such as data quality checks or producing metrics to summarize model performance, as in model diagnostics. For example, the `assertr` package [@assertr] and the `pointblank` package [@pointblank] provide data validation tools to allow users to define data quality checks. In contrast, packages provide model checks tools like `performance` [@performance] and `yardstick` [@yardstick], from the `tidymodels` [@tidymodels] ecosystem, offers goodness-of-fit, residual analysis, and model performance metrics.

These packages provide the tools to conduct diagnostics but still don't reflect the mental process of how data analysts diagnose the unexpected output. For example, when an unexpected output occurs, an analyst may check on whether a column in the data frame is between two values for data quality. However, it is not documented what motivates the analyst to conduct this check -- whether it also applies to other researchers analyzing new data in the same contexts, whether it is a common practice in the field, or whether it is a reaction to this particular data or scenario. Currently, most of these assumptions are not discussed in the publication or captured by tools themselves. While one might be able to infer some of the mental process of the analysts from external sources, such as talking to them or watching screencast videos produced by the analysts e.g. TidyTuesday screencast videos, these insights are not systematically documented or made machine-readable. This gap highlights the need for tools that provide higher level documentation of reasoning behind the checks, facilitating a more transparent and interpretable analysis process.
